Quantifying Stability and Change in Personal Culture Using Panel Data

### 2023-09-29 manuscript/manuscript.tex
small update in the manuscript.
SV edits

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items preferred the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they showed no evidence of durable
change. Questions with more evidence of durable change included salient
issues like gay marriage and questions tapping ``public'' statements or
behaviors such as partisan identification and religious service
attendance. There was also more evidence of durable change among early
adults (people ages 18-30) than the rest of the population. Overall, the
researchers concluded that ``results ultimately suggest that real,
persistent attitude change is an uncommon phenomenon among adults'\,'
(Kiley and Vaisey 2020: p.~500; see also Vaisey and Kiley 2021). This
lack of durable change is consistent with other recent findings that
cohort replacement plays a larger role than period effects in explaining
differences in personal culture (Vaisey and Lizardo 2016).

@@

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM does this by modeling each individual as
following their own linear trajectory. But rather than debating whether
this model fits a particular data set better than a model that assumes
that people never change, researchers should quantify these individual
changes and compare those to the differences that exist between people
at the panel's start. In other words, the focus should shift from
asking, ``do people change?'' to asking, ``what are the relative
contributions of change and initial differences for explaining
differences in personal culture?''

We propose an approach that offers such a quantification. In this
approach, there are two metrics that each reflect a combination of
relevant theoretical processes. The first metric, interpersonal
differences at the start of the survey, reflects the accumulated
experiences of people prior to entering the panel survey that shape how
they answer questions. Lersch calls this ``early imprinting.'' While
commonly associated with experiences during a formative period that
result in settled dispositions, this variation could also reflect
experiences that happened at any stage in life as long as they predate
the panel and consistently affect subsequent responses. For instance,
for those people entering the panel post-retirement, this ``imprinting''
might reflect this pivotal life transition. Consequently, this metric
also reflects variation in individuals' social roles or statuses at the
start of the panel that were important in shaping their dispositions.

The second metric, the amount of intrapersonal linear change people make
during the panel, captures durable changes in personal culture over
time, a set of processes collectively called ``persistent change'' or
``adaption'' by Lersch and ``active updating'' by Kiley and Vaisey.
Lersch attributes these changes to social triggers such as moving into a
new environment or adopting new social roles (although he never measures
them directly). Additionally, they might reflect the diffusion of new
cultural forms across social networks, cues from political elites or
otherwise culturally influential leaders, the emergence of issues in
politics or culture, or large-scale social shifts.

A third metric, often called ``residual variance'' or ``fluctuation,''
accounts for the remainder of variance in peoples' responses. This
non-durable change emerges for a variety of reasons. For example, people
might not have a clear disposition on a particular item as it is asked.
Instead, they might internalize a broad set of considerations and
construct an opinion in the context of the survey interview, with
different considerations coming to the forefront of their cognition
during each interview (Feldman and Zaller 1992; Tourangeau, Rips, and
Rasinski 2000; Zaller 1992). Additionally, this variance can also
include measurement error, such as misinterpreted questions, erroneous
response selections, or responses getting coded incorrectly. While this
third metric at least partially reflects important processes of personal
culture, it does not directly touch on the ongoing debate outlined here.
Because of this we quantify this component but focus principally on the
other two.

@@

By shifting the question and adjusting the approach, we expect
consistent patterns across datasets previously analysed in the debate.
First, because previous work found evidence of durable change on most
measures of personal culture, we expect that intrapersonal change will
have some explanatory power for most items. But because these earlier
studies also found many questions (between 25 and 35 percent) lacking
evidence of durable change, we expect considerable variation in the
variance explained by intrapersonal change, with some questions
indicating essentially no variance attributable to such change.

Second, we expect the variance explained by intrapersonal change to be
modest compared to interpersonal differences at the start of the survey.
This expectation should be uncontroversial, regardless of the
theoretical process assumed to be most relevant for a particular
question. Panel surveys do not encompass individuals' entire lifetimes.
Thus, even if intrapersonal change during adulthood is the major driver
of cultural differences, the change detected in a panel covering 20
years probably will not fully account for the entire variance of
interpersonal differences.

Our expectations so far largely echo what was observed in previous
studies. However, theoretical perspectives differ in how much variance
they would expect to be driven by intrapersonal change, enabling
quantification to provide a new perspective on existing questions.
``Formative period'' and cohort theories, which emphasize early-life
socialization as a strong force in shaping adult personal culture,
suggest that the variance from intrapersonal change would still fall
short of explaining interpersonal differences, even when we account for
the duration of observation. Conversely, other theoretical perspectives
emphasizing the importance of people's contemporaneous social
environment for personal culture expect intrapersonal change to account
for a large portion of the variance, regardless of the duration.

@@

We use data from seven nationally representative panel surveys from
Australia, Germany, Great Britain, Switzerland, and the United States
(summarized in Table 1), combining all the data files used in previous
work (Kiley and Vaisey 2020; Lersch 2023).\footnote{For more information
  on these data sources, see (Goebel et al. 2019; Income Dynamics 2013;
  Smith et al. 2022; Summerfield et al. 2011; Taylor 1996; University of
  Essex and Research 2019; Voorpostel et al. 2016).} These studies cover
a long period of time (the range of surveys spans from 1968 to 2021),
with 609 personal culture items that capture attitudes, beliefs, values,
self-assessments, self-descriptions, and behaviors (Alwin 2007). We
restricted the sample such that individuals between the ages of 18 and
79 are included without further elimination, and in all surveys, we used
all possible cases for which respondents provided responses. In the end,
the analyses that follow rely on a cross-cultural sample with a
cross-domain set of items to capture a broad range of individual
personal culture. Supplemental Materials A documents the list of all
variables used in the upcoming analyses.

@@

We start with Lersch's (2023) Life Course Adaption Model, which
formalizes survey responses at time \(t\) as a function of
individual-level random intercepts and slopes for survey age, what is
often called a mixed-effects growth curve model. This model formally
assumes a set of propositions about change that reflect the theoretical
debate to this point. First, consistent with the settled disposition
model, it assumes that people start the survey with cultural
differences, modeled as random intercepts for each respondent. Second,
it assumes that people change over time, taking the form of random
slopes for each respondent as a linear function of time. Third, it
assumes that people deviate around this baseline randomly over time,
reflecting ``fluctuation'' or short-term non-persistent change. Finally,
it assumes that people draw on earlier time points in reacting to new
events. Formally, this can be written as

\[
y_{i, t} = \beta_0 + u_{0, i} + \beta_{1} x_t + u_{1, i} x_{i, t} + \epsilon_{i, t}
\] \[
u_{0,i} \sim \mathcal{N}(0,\sigma_{\mu_0}^2) \\
\] \[
u_{1,i} \sim \mathcal{N}(0,\sigma_{u_1}^2)
\]

where \(\beta_0\) is the average intercept, \(u_{0, i}\) is the random
intercept for individuals, \(\beta_1 x_t\) is the average slope,
\(u_{1, i} x_{i, t}\) is the random slope for individuals, and
\(\epsilon_{i, t}\) is the random error term that captures transitory
disturbances.

@@

As a first step in our analysis, we fit a LCAM to each of the 609
measures of personal culture outlined above. We then use these models to
derive two measures of systematic variance that correspond to the two
collections of theoretical processes outlined above: interpersonal
differences at baseline and intrapersonal change over time. First, to
measure the importance of interpersonal differences, we calculate the
sum of the squared residuals derived from measuring each respondent at
the midpoint of their age trajectory.\footnote{We use the scale
  midpoint, rather than the actual difference at baseline, to better
  capture the differences between people independent of their age
  trajectories.} We then divide this value by the total sum of squares
and subtract it from from 1, producing a value we denote as V(D), or the
proportion of variance attributable to interpersonal differences. This
number is equivalent to an \(R^2\) value generated by using random
intercepts for all respondents. Second, we calculate the sum of squared
residuals generated by additionally accounting for the age trajectory.
We divide this number by the total sum of squares and subtract it from
1. We then subtract \(V(D)\) from this value. This gives us a value we
denote as \(V(C)\), the additional proportion of variance explained by
intrapersonal change.

These two numbers reflect the total systematic variance of the model
that can be attributed to either differences across people or linear
change over the course of the panel.\footnote{This does not mean that
  other factors might not explain additional variance if they were
  included in the model. For example, peoples' responses about religious
  beliefs and participation might systematically vary by the day of the
  week or month in which they are surveyed. This variation is moved into
  the ``random'' or non-systematic component of the model for us.} Our
principal outcome of interest is the proportion of systematic variance
accounted for by intrapersonal change, which we calculate as

@@

We can think of this proportion \(\omega\) as akin to an Intraclass
Correlation Coefficient focused only on the systematic components of the
model while ignoring residual variance that might be attributable to
measurement error or non-persisting change (see also the variance
decomposition of period and cohort by Vaisey and Lizardo 2016). The
higher this proportion is, the more systematic variance in responses is
attributable to intrapersonal change during the panel. The lower this
proportion, the less intrapersonal change seems to play a large role in
explaining systematic cultural differences during the panel. Moving
forward, we use \(\omega\) as a placeholder for \emph{systematic
variance attributable to intrapersonal change}.

Because this quotient summarizes variance and therefore has no natural
referent, it is hard to say at which values we observe ``a lot'' of
intrapersonal change. But summarizing questions in this way allows us to
compare the relative importance of our two broad processes across a
range of questions that might have different levels of non-systematic
variance, giving us a sense of the relative prevalence of these two
processes for each question (something not achievable under the
tournament of models approach), and allowing us to compare which kinds
of questions show more or less intrapersonal change relative to
interpersonal differences. While the tournament of models approach
allowed researchers to say whether questions in general showed evidence
of updating, this approach allows us to make more specific claims about
the prevalence of these theoretical processes for particular questions
relative to each other.

As a second step in our analysis, we use ordinary least squares
regression to model \(\omega\) as a function of features of the question
and panel. This allows us to explore how these features affect the
prevalence of intrapersonal change. These covariates include the number
of response options (measured as 2, 3-5, and more than 5), the panel,
the log of the number of participants, the number of waves observed, the
time since the question was first asked, and the total duration of time
the question was asked.

@@

Figure 1 plots the proportion of systematic variance attributable to
interpersonal differences and intrapersonal change for each of the 609
questions analyzed, plotted separately by panel. To broadly summarize
the results, all panels show a similar range of the proportion of
systematic variance attributable to intrapersonal change. Across panels,
the median value of \(\omega\) is 0.094 (mean of 0.098), with an
interquartile range of 0.053 to 0.139. All panels include questions
where \(\omega\) is essentially 0, and all panels include questions with
\(\omega\) values greater than 0.20. The maximum \(\omega\) value is
0.26.

@@

Across all questions, interpersonal differences explain a much larger
share of the systematic variance in responses than intrapersonal change.
Again, this is to be expected. Interpersonal differences capture not
just pre-adult socialization, but all accumulated experiences up to the
start of the panel that might have a bearing on personal culture.

To the extent that there are differences across the panels, the PSID has
the highest \(\omega\) values with mean .120 and median .135. While we
cannot disentangle features of the sample from features of the questions
asked to each sample, the specific samples for many PSID questions have
lower average ages than those from other panels. To the extent that
younger respondents might be more likely to make durable changes of
opinion, this higher explanatory power of intrapersonal change might
reflect the distinct age profile of respondents in this sample. At the
other end, the GSS has the lowest range of \(\omega\) with mean .073 and
median .069. This potentially reflects the fact that the GSS observes
people for a shorter duration, on average, than the other panels. If,
consistent with life course adaption theories, people are more likely to
make significant cultural changes the longer we observe them, then
duration likely affects the range of \(\omega\) (a point we explore more
below). However, the GSS results are still consistent with results from
the other panels.

While there are some differences between panels, these differences are
small compared to the differences within panels. For about 6 percent of
items explored here, \(\omega\) is greater than 0.20. These questions
tend to ask about external referents (e.g., confidence in specific
government leaders or political parties), life satisfaction, or current
financial position. At the other end, questions about religious
identification, views on gender roles, and support for civil liberties
tend to have very low proportion of variance attributable to
intrapersonal change.

In contrast to the tournament of models approach, quantifying change
this way allows us to explore variation in the relative importance of
intrapersonal change across questions that all show evidence of change.
For example, Kiley and Vaisey (2020) found that confidence in the press
and confidence in religious leaders were both characterized by active
updating. Our results show that intrapersonal change is much more
important for explaining variance in confidence in the press (0.164)
than confidence in religion (0.049).

Supplemental Materials A presents the estimated proportion of variance
attributable to interpersonal differences and intrapersonal change, the
estimated values of \(\omega\), and the proportion of residual variance
for each question analyzed. In general, interpersonal differences are
almost always the largest component of the total variance and tend to
explain between 55 and 70 percent of total variance, while intrapersonal
change is always the smallest, typically explaining between 3 and 8
percent of total variance. Residual variance tends to account for
between 22 and 37 percent of variance, though on several questions
residual variance is greater than 50 percent.

As noted above, the substantive importance of intrapersonal change for
understanding cultural differences depends on a range of factors,
including how long the panel runs, whether assumptions about linear
change hold, and whether the panel is capturing a distinctly turbulent
period or a distinctly stable period for the item under question.
However, if we assume that the period under observation is ``typical''
for a question ---not a time of extremely heightened (or lowered)
sensitivity or change--- then it does not seem realistic for this
process of intrapersonal change to explain a large share of
interpersonal differences for many of the questions we observe.

@@

Figure 2 plots the results from an ordinary least squares regression of
\(\omega\) as a function of question, panel, and sample features. These
models also include fixed effects for panels and topics, so coefficients
reflect the association within a panel and topical domain.

@@

While we did not generate expectations for how question content relates
to \(\omega\), some of the associations between question structure,
panel duration, and panels themselves might be driven by differences in
the topics addressed by each panel. To address this, we followed Hout et
al. (2016) and Lersch (2023) in coding each question as falling into one
of nine different content categories and included these as indicator
variables in the meta analysis. Supplemental Materials B presents the
estimates for these content indicators. They show some associations with
\(\omega\), with questions about subjective SES; social life, social
cohesion, and trust; environment and climate; and health and morale
showing larger \(\omega\) values than questions about religion and
spirituality; politics, government and the economy; and gender and
family.

The most notable result in the meta analysis is the negative coefficient
attached to the duration of years covered by the question. The longer a
question is observed, the less is the value of \(\omega\). Theories of
personal cultural change that link changes in personal culture to social
experiences, including the LCAM, suggest that the longer we observe
respondents, the more likely people are to undergo potentially
transformative experiences and therefore the more variance would be
explained by intrapersonal change. The fact that we find a negative
coefficient attached to this variable can be a challenge to that
assumption.

As a further test of this coefficient, we compared the \(\omega\) values
when using the full duration of a panel compared to when we dropped the
final wave and therefore reduced the total duration of observation for
the question. If the coefficient reflects a true negative effect of
duration on \(\omega\), we should see that same effect within questions.
Results from this analysis are presented in Supplemental Materials C and
contradict the coefficient from the regression model: we found that the
longer we observe the same question, the higher the \(\omega\) is. We
interpret this combination of findings as suggesting that the kinds of
questions asked for a longer time period tend to demonstrate less
intrapersonal change than questions asked over shorter periods, rather
than a true function of time. This reinforces the explanation for why
the GSS displays the lowest average proportion of systematic variance
attributable to intrapersonal change: its shorter duration.

@@

This work has tended to focus on the fact that college-educated
Americans give responses that are less likely to be affected by
measurement error or short-term influences than the rest of the
population (Alwin 2007; Zaller 1992). But focus on the difference in the
non-systematic or residual component of variance across groups obscures
the fact that the relative importance of different components of
systematic variance, intrapersonal change and interpersonal differences,
might also differ across these groups.

There are theoretical reasons to believe that education might be
associated with either more or less systematic variance attributable to
intrapersonal change. On one hand, because college graduates are more
connected to mainstream discourse and elite signals, they might be more
likely to make durable changes in response to the emergence of new
information, new issues, or political realignments, while those without
chronic information display more variance around their baselines without
their baselines changing much (Zaller 1992). In contrast, it could be
the case that the stability demonstrated by educated respondents
reflects the fact that they have already formed durable opinions and are
relatively closed off to new information. In this perspective, college
could serve as a formative experience that solidifies some dimensions of
personal culture for people who attend. But among people who do not
attend college, later life experiences might prove more important in
forming or changing personal culture, as these experiences potentially
provide information that college-educated peers received earlier.

To test these competing propositions, we use the models presented above
to separately calculate \(\omega\) values for people with at least a
bachelor's degree and people with less than a bachelor's degree at wave
1\footnote{A small number of respondents report different highest
  degrees at wave 1, wave 2, and wave 3. Some of this is due to
  measurement error, and some of it is due to a small number of people
  obtaining a higher degree during the four years of the panel.
  Estimating the panel with highest at wave 1 or highest degree reported
  across the panel produces functionally identical results.} of the
General Social Survey's panels. We focus on the GSS because it contains
the largest number of questions tapping general political dispositions,
which is the domain where education has proven particularly relevant for
understanding attitude stability. The GSS also covers a turbulent window
of American politics from 2006 to 2014. This window covers the start of
the Great Recession, debates about federal intervention in and
regulation of Wall Street, the election of Barack Obama as the first
black president in U.S. history, debates about the role of the federal
government in the health care sector, the emergence of the Tea Party,
and political realignment and clarification on the issue of gay
marriage, among other topics.

Figure 3 plots the distribution of differences in \(\omega\) values
between people with at least a bachelor's degree and people with less
than a bachelor's degree at wave 1 of the panel for all 183 GSS items we
explored. Values greater than 0 indicate that intrapersonal change is
more important in explaining the responses of college graduates than
those of people without a college degree, while values less than 0
indicate that intrapersonal change is less important for explaining
variation among people with a college degree than those without a
college degree.

>>>>

Empirically comparing which of these models better fit a broad range of
questions from the General Social Survey's rotating panels, Kiley and
Vaisey (2020) found limited evidence of durable change. While the
majority of items preferred the AUM, the amount of durable change
detected on these items was small. A substantial minority of questions
(39 percent) favored the SDM, meaning they showed no evidence of durable
change. Questions with more evidence of durable change included salient
issues like gay marriage and questions tapping ``public'' statements or
behaviors such as partisan identification and religious service
attendance. There was also more evidence of durable change among early
adults (people ages 18-30) than the rest of the population. Overall, the
researchers concluded that ``results ultimately suggest that real,
persistent attitude change is an uncommon phenomenon among adults'\,'
(Kiley and Vaisey 2020: p.~500; see also Vaisey and Kiley 2021). This
lack of durable change is consistent with other recent findings that
cohort replacement plays a somewhat larger role than period effects in
explaining differences in personal culture (Vaisey and Lizardo 2016).

@@

A more theoretically productive approach begins with a model assuming
that during an observed time period people might remain stable, might
change a little, or might undergo significant shifts in their personal
culture. Lersch's LCAM does this by modeling each individual as
following a personal linear trajectory. But rather than debating whether
this model fits a particular data set better than a model that assumes
that people never change, researchers should quantify these individual
changes and compare those to the differences that exist between people
at the panel's start. In other words, the focus should shift from
asking, ``do people change?'' to asking, ``what are the relative
contributions of change and initial differences for explaining variance
in personal culture?''

We propose an approach that offers such a quantification. Our measure
relates two metrics that each reflect a combination of relevant
theoretical processes. The first metric, interpersonal differences at
the start of the survey, reflects the accumulated experiences of people
prior to entering the panel survey that shape how they answer questions.
Lersch calls this ``early imprinting.'' While commonly associated with
experiences during a formative period that result in settled
dispositions, this variation could also reflect experiences that
happened at any stage in life as long as they predate the panel and
consistently affect subsequent responses. For instance, for those people
entering the panel post-retirement, this ``imprinting'' might reflect
this pivotal life transition. Consequently, this metric also reflects
variation in individuals' social roles or statuses at the start of the
panel that were important in shaping their dispositions.

The second metric, the amount of intrapersonal linear change people make
during the panel, captures durable changes in personal culture over
time. This captures the set of processes collectively called
``persistent change'' or ``adaption'' by Lersch and ``active updating''
by Kiley and Vaisey. Lersch attributes these changes to social triggers
such as moving into a new environment or adopting new social roles
(although he never measures these directly). Additionally, they might
reflect the diffusion of new cultural forms across social networks, cues
from political elites or otherwise culturally influential leaders, the
emergence of issues in politics or culture, or large-scale social
shifts.

A third metric, which we might call ``residual variance'' or
``fluctuation,'' accounts for the remainder of variance in peoples'
responses. These non-durable changes emerge for a variety of reasons.
For example, people might not have a clear disposition on a particular
item as it is asked. Instead, they might internalize a broad set of
considerations and construct an opinion in the context of the survey
interview, with different considerations coming to the forefront of
their cognition during each interview (Feldman and Zaller 1992;
Tourangeau, Rips, and Rasinski 2000; Zaller 1992). This variance can
also include measurement error, such as misinterpreted questions,
erroneous response selections, or responses getting coded incorrectly.
Although this third metric at least partially reflects important
processes of personal culture, it does not directly touch on the ongoing
debate outlined here. Because of this, we quantify this component but
focus principally on the other two.

@@

By shifting the question and adjusting the approach, we expect
consistent patterns across datasets previously analysed in the debate.
First, because previous work found evidence of durable change on most
measures of personal culture, we expect that we will find some
intrapersonal change in most items. But because these earlier studies
also found many questions (between 25 and 35 percent) lacking evidence
of \emph{any} durable change, we expect considerable variation in the
variance accounted for by intrapersonal change, with some questions
indicating essentially no variance attributable to such change.

Second, we expect the variance attributable to intrapersonal change to
be modest compared to interpersonal differences at the start of the
survey. This expectation should be uncontroversial, regardless of the
theoretical process assumed to be most relevant for a particular
question. Panel surveys do not encompass individuals' entire lifetimes.
Thus, even if intrapersonal change during adulthood is the major driver
of cultural differences, the change detected in a panel covering 20
years probably will not fully account for the entire variance of
interpersonal differences.

Our expectations so far largely echo what was observed in previous
studies. However, theoretical perspectives differ in how much variance
they would expect to be driven by intrapersonal change, enabling
quantification to provide a new perspective on existing questions.
``Formative period'' and cohort theories, which emphasize early-life
socialization as a strong influence on adult personal culture, suggest
that the variance from intrapersonal change would still fall short of
explaining interpersonal differences, even when we account for the
duration of observation. Conversely, other theoretical perspectives
emphasizing the importance of people's contemporaneous social
environment for personal culture might expect intrapersonal change to
account for a large portion of the variance, regardless of the duration.

@@

We use data from seven nationally representative panel surveys from
Australia, Germany, Great Britain, Switzerland, and the United States
(summarized in Table 1), combining all the data files used in previous
work (Kiley and Vaisey 2020; Lersch 2023).\footnote{For more information
  on these data sources, see (Goebel et al. 2019; Income Dynamics 2013;
  Smith et al. 2022; Summerfield et al. 2011; Taylor 1996; University of
  Essex and Research 2019; Voorpostel et al. 2016).} These studies cover
a long period of time (the range of surveys spans from 1968 to 2021),
with 609 personal culture items that capture attitudes, beliefs, values,
self-assessments, self-descriptions, and behaviors (Alwin 2007). We
restricted the sample such that individuals between the ages of 18 and
79 are included without further elimination, and in all surveys, we used
all possible cases for which respondents provided responses. In the end,
the analyses that follow rely on a cross-national sample with a
cross-domain set of items to capture a broad range of individual
personal culture. Supplemental Materials A documents the list of all
variables used in the upcoming analyses.

@@

We start with Lersch's (2023) Life Course Adaption Model, which
formalizes survey responses at time \(t\) as a function of
individual-level random intercepts and slopes for survey age, what is
often called a mixed-effects growth curve model. This model formally
assumes a set of propositions about change that reflect the theoretical
debate to this point. First, consistent with the settled disposition
model, it assumes that people start the survey with cultural
differences, modeled as random intercepts for each respondent. Second,
it assumes that people change over time, taking the form of random
slopes for each respondent as a linear function of time. Third, it
assumes that people deviate around this baseline randomly over time,
reflecting ``fluctuation'' or short-term non-persistent change.
Formally, this can be written as

\[
y_{it} = \beta_0 + \alpha_{0i} + (\beta_1 + \alpha_{1i}) \text{year}_{it} + \epsilon_{it}
\]

\[
\begin{aligned}
\alpha_{0i} &\sim \mathcal{N}(0,\tau^2_{0}) \\
\alpha_{1i} &\sim \mathcal{N}(0,\tau^2_{1}) \\
\epsilon_{it} &\sim \mathcal{N}(0,\sigma^2)
\end{aligned}
\]

where \(\beta_0\) is the average intercept, \(\alpha_{0i}\) is the
random intercept component for individuals, \(\beta_1\) is the average
yearly change in the outcome, \(\alpha_{1i}\) is the random slope
component for individuals, and \(\epsilon_{it}\) is the random error
term that captures transitory disturbances. The \(\alpha_{0i}\) and
\(\alpha_{1i}\) terms are also allowed to covary.

@@

As a first step in our analysis, we fit this LCAM to each of the 609
measures of personal culture outlined above. We then use these models to
derive two measures of systematic variance that correspond to the two
collections of theoretical processes outlined above: interpersonal
differences at baseline and intrapersonal change over time.

We estimate these using a form of variance decomposition. First, to
measure stable interpersonal differences, we calculate the sum of the
squared residuals derived from assuming that each person spends the
entire survey at the estimated midpoint of their LCAM
trajectory.\footnote{We use the midpoint estimate, rather than the value
  at the first wave, because this is the best measure of ``baseline''
  available under the assumption that the person does not change. The
  first wave measure alone contains an unknown amount of measurement
  error and transient fluctuations.} We then divide this value by the
total sum of squares and subtract this number from from 1, producing a
value we denote as \(V(D)\). This is the proportion of total variance
attributable to interpersonal differences. (This number is equivalent to
the conditional \(R^2\) value generated by a model with only random
intercepts for all respondents.) Second, we calculate another sum of
squared residuals generated by allowing the predictions to vary by
estimated LCAM trajectory. We divide this number by the total sum of
squares and subtract it from 1. We then subtract \(V(D)\) from this
value. This gives us a value we denote as \(V(C)\), the additional
proportion of variance accounted for by intrapersonal change.

These two numbers together reflect the total systematic variance of the
LCAM that can be attributed to either differences between people or
linear intrapersonal change over the course of the panel.\footnote{This
  does not mean that other factors might not explain additional variance
  if they were included in the model. For example, peoples' responses
  about religious beliefs and participation might systematically vary by
  the day of the week or month in which they are surveyed. This
  variation is moved into the ``random'' or non-systematic component of
  the model for us.} Our principal outcome of interest is the proportion
of systematic variance accounted for by intrapersonal change, which we
calculate as

@@

We can think of \(\omega\) as a type of Intraclass Correlation
Coefficient focused only on the systematic components of the model while
ignoring residual variance that might be attributable to measurement
error or non-persisting change (see also the variance decomposition of
period and cohort by Vaisey and Lizardo 2016). The higher this
proportion is, the more systematic variance in responses is attributable
to intrapersonal change during the panel. From now on, we use \(\omega\)
as a placeholder for \emph{systematic variance attributable to
intrapersonal change}.

Because this quotient summarizes variance and therefore has no natural
referent, it is hard to say at which values we observe ``a lot'' of
intrapersonal change. But summarizing questions in this way allows us to
compare the relative importance of our two broad processes across a
range of questions that might have different levels of non-systematic
variance. This gives us a sense of the relative prevalence of these two
processes for each question, which is not achievable under the
tournament of models approach. This also allows us to compare which
kinds of questions show more or less intrapersonal change relative to
interpersonal differences. While the tournament of models approach
allowed researchers to say whether questions in general showed
\emph{any} evidence of updating, this approach allows us to make more
specific claims about the prevalence of these theoretical processes for
particular questions relative to each other.

As a second step in our analysis, we use linear regression to conduct a
meta-analysis, modeling \(\omega\) as a function of features of the
question and panel. This allows us to explore how these features are
associated with the degree of intrapersonal change. These covariates
include the number of response options (measured as 2, 3-5, and more
than 5), the specific survey, the log of the number of participants, the
number of waves observed, the time period (i.e., decade) the question
was first asked, and the total duration of time the question was asked.

@@

Figure 1 plots the proportion of systematic variance attributable to
interpersonal differences and intrapersonal change for each of the 609
questions, plotted separately by panel. To broadly summarize the
results, all panels show a similar range of the proportion of systematic
variance attributable to intrapersonal change. Across panels, the median
value of \(\omega\) is 0.094 (mean of 0.098), with an interquartile
range of 0.053 to 0.139. All panels include questions where \(\omega\)
is essentially 0, and all panels include questions with \(\omega\)
values greater than 0.20. The maximum \(\omega\) value is 0.26.

@@

Across all questions, interpersonal differences account for a much
larger share of the systematic variance in responses than intrapersonal
change. Again, this is to be expected. Interpersonal differences capture
not just pre-adult socialization, but all accumulated experiences up to
the start of the panel that might influence personal culture.

To the extent that there are differences across the panels, the PSID has
the highest \(\omega\) values with mean .120 and median .135. While we
cannot disentangle features of the sample from features of the questions
asked to each sample, the specific samples for many PSID questions have
lower average ages than those from other panels. To the extent that
younger respondents might be more likely to make durable changes of
opinion, these higher estimates of intrapersonal change might reflect
the distinct age profile of respondents in this sample. At the other
end, the GSS has the lowest range of \(\omega\) with mean .073 and
median .069. This potentially reflects the fact that the GSS observes
people for a shorter duration, on average, than the other panels. If,
consistent with life course adaption theories, people are more likely to
make significant cultural changes the longer we observe them, then
duration likely affects the range of \(\omega\) (a point we explore more
below). However, the GSS results are still consistent with results from
the other panels.

While there are some differences between panels, these differences are
small compared to the differences within panels. For about 6 percent of
items, \(\omega\) is greater than 0.20. These questions tend to ask
about objectively changing external referents (e.g., confidence in
specific government leaders or political parties), life satisfaction, or
current financial position. At the other end, questions about religious
identification, views on gender roles, and support for civil liberties
tend to have very low estimates of intrapersonal change.

In contrast to the tournament of models approach, quantifying change
this way allows us to explore variation in the relative importance of
intrapersonal change across questions that all show evidence of change.
For example, Kiley and Vaisey (2020) found that confidence in the press
and confidence in religious leaders were both characterized by active
updating. Our results show that intrapersonal change is much more
important for explaining variance in confidence in the press (0.164)
than confidence in religion (0.049), even though both are updating.

Supplemental Materials A presents the estimated proportion of variance
attributable to interpersonal differences and intrapersonal change, the
estimated values of \(\omega\), and the proportion of residual variance
for each question. Interpersonal differences are almost always the
largest component of the total variance and tend to account for between
55 and 70 percent of total variance, while intrapersonal change is
always the smallest, typically accounting for between 3 and 8 percent of
total variance. Residual variance tends to account for between 22 and 37
percent of variance, though on several questions residual variance is
greater than 50 percent. This might indicate survey items with low
reliability or ones that capture genuinely rapid fluctuations.

As we noted above, the substantive importance of a given amount of
intrapersonal change depends on a range of factors, including how long
the panel runs, whether assumptions about linear change hold, and
whether the panel is capturing a distinctly turbulent period or a
distinctly stable period for the relevant item. However, if we assume
that the period under observation is ``typical'' for a question---not a
time of extremely heightened (or lowered) sensitivity or change---then
it does not seem realistic for that intrapersonal change accounts for a
large share of the cultural differences we see in the world.

@@

Figure 2 plots the results from a linear regression of \(\omega\) as a
function of question, panel, and sample features. These models also
include fixed effects for panels and topics, so coefficients reflect the
association within a panel and topical domain.

@@

Although we did not state expectations for how question content would
relate to \(\omega\), some of the associations between question
structure, panel duration, and panels themselves might be driven by
differences in the topics addressed by each panel. To address this, we
followed Hout et al. (2016) and Lersch (2023) in coding each question as
falling into one of nine different content categories and included these
as indicator variables in the meta analysis. Supplemental Materials B
presents the estimates for these content indicators. They show some
associations with \(\omega\), with questions about subjective SES;
social life, social cohesion, and trust; environment and climate; and
health and morale showing larger \(\omega\) values than questions about
religion and spirituality; politics, government and the economy; and
gender and family.

The most notable result in the meta analysis is the negative coefficient
attached to the duration of years covered by the question. The longer a
question is observed, the less is the value of \(\omega\). Theories of
personal cultural change that link changes in personal culture to social
experiences, including the LCAM, suggest that the longer we observe
respondents, the more likely people are to undergo potentially
transformative experiences and therefore the more variance would be
attributable to intrapersonal change. Finding a negative coefficient
here seems to challenge to that assumption.

As a further test of this finding, we compared the \(\omega\) values
when using the full duration of a panel compared to when we dropped the
final wave and therefore reduced the total duration of observation for
the question. If the coefficient reflects a true negative effect of
duration on \(\omega\), we should see that same effect within questions.
Results from this analysis are presented in Supplemental Materials C and
contradict the coefficient from the regression model; with a few
exceptions, we found that the longer we observe the same question, the
higher \(\omega\) is. We interpret this combination of findings as
suggesting that the kinds of questions asked for a longer time period
tend to demonstrate less intrapersonal change than questions asked over
shorter periods, rather than a true function of time. This supports the
conclusion that the GSS shows less intrapersonal change because of its
shorter duration.

@@

This work has tended to focus on the fact that college-educated
Americans give responses that are less likely to be affected by
measurement error or short-term influences than the rest of the
population (Alwin 2007; Zaller 1992). But this focus on the
non-systematic or residual component of variance across groups lumps the
two systematic forms of difference together. That is, it obscures the
fact that the amount of interpersonal differences and intrapersonal
change might also differ across these groups.

There are theoretical reasons to believe that education might be
associated with either more or less intrapersonal change. On one hand,
because college graduates are more connected to mainstream discourse and
elite signals, they might be more likely to make durable changes in
response to the emergence of new information, new issues, or political
realignments, while those without chronic information might display more
variance around an unchanging baseline (Zaller 1992). Conversely, it
could be that the observed stability of the college educated reflects
the fact that they have already formed durable opinions and are
relatively closed off to new information. If this is true, college could
be understood as a formative experience that solidifies some dimensions
of personal culture. Perhaps for those who do not attend college, later
life experiences might prove more important in forming or changing
personal culture, as these experiences potentially provide information
that college-educated peers have already received.

To compare these competing propositions, we calculate \(\omega\) values
separately for people with at least a bachelor's degree and people with
less than a bachelor's degree at wave 1\footnote{A small number of
  respondents report different highest degrees at wave 1, wave 2, and
  wave 3. Some of this is due to measurement error, and some of it is
  due to a small number of people obtaining a higher degree during the
  four years of the panel. Estimating the panel with highest at wave 1
  or highest degree reported across the panel produces functionally
  identical results.} of the three General Social Survey's panels. We
focus on the GSS because it contains the largest number of questions
tapping general political dispositions, which is the domain where
education has proven particularly relevant for understanding attitude
stability. The GSS also covers a turbulent window of American politics
from 2006 to 2014. This window covers the start of the Great Recession,
debates about federal intervention in and regulation of Wall Street, the
election of Barack Obama as the first black U.S. president, debates
about the role of the federal government in the health care sector, the
emergence of the Tea Party, and political realignment and clarification
on the issue of gay marriage, among other topics.

Figure 3 plots the distribution of differences in \(\omega\) values
between people with at least a bachelor's degree and people with less
than a bachelor's degree at wave 1 of the panel for 183 GSS items.
Values greater than 0 indicate that intrapersonal change accounts for
more systematic variance among college graduates than among those
without a college degree, while values less than 0 indicate the
opposite.
